Leadership across cultures and contexts: an exploratory analysis

### 2020-06-24 leadership_across_cultures_contexts.Rmd
Spell check.

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
<!-- Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group contex. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2020; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2020; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2020; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020]. -->

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, community leadership is typically informal and emerges in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015]. Leaders also play important roles in ritual and religeous contexts [@singh_cultural_2017; @winkelman_crosscultural_2020], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Leadership in between-group conflict and cooperation is also common across nonindustrial societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community, and institutionalized managerial roles are associated with sedentarization, defendable resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of resources creates increased opportunities for resource management and is often associated with increased territoriality and inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with privlaged social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize political influence and maintain control through a variety of mechanisms including military force, ideologies, and control of information [@bodley_cultural_2011].

@@

Subsistence strategy, group structure and continental region were confounded. Hunter-gatherer text records often referred to residential and kin leadership, but rarely to supracommunity leadership; agriculturalist text records rarely referred to residential leadership but often referred to supracommunity leadership. Community-level political groups were frequent across all subsistence types. African text records were predominantly from horticulturalits whereas North American text records were predominantly from hunter-gatherers See Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence), and Table S\@ref(tab:fourfactorsumm).

@@

There were no text records that described shamans or leaders with supernatural qualities who were exclusively female. However, there were `r sum(df_shaman$demo_sex == 'both' & df_shaman$shamanism)` text records describing such leaders of both sexes. There was also more evidence of shamans and leaders with supernatual qualities in the Americas compared to other regions. See Figure S\@ref(fig:shamanregion).

@@

The leadership features we derived by clustering leadership dimensions were represented across most cultures. Leaders performed three broad functions: providing prosocial services, strategizing, and organizing collective actions. Regarding characteristic traits, leaders possessed some combination of individual competencies, many of which are prosocial, conformed to cultural norms, and receivied various benefits (Figure \@ref(fig:featuressupportplot)).

@@

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (Figure \@ref(fig:varsupportcb)). Nevertheless, for most dimensions of benefits and costs, there was more evidence of benefits than costs (Figure \@ref(fig:benefitcostratio)), and the same was true averaging across all benefit/cost dimensions ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$). Additionally, several positive leader qualities indicating social, reproductive, and material success clustered together (bottom cluster Figure \@ref(fig:cluster)A). These results are consistent with much theoretical literature that links leader benefits to their capacities to provide prosocial benefits (e.g., resolving conflicts, organizing collective actions, and generousity) and/or impose social costs (e.g., punishing and aggressiveness) [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013; @hooper_theory_2010; @gavrilets_solution_2014; @price_evolution_2014; @smith_leadership_2016].

@@

The potential evolutionary theoretical importance of shaman leaders (which in our data includes all leaders with supernatural qualities), highlights the importance of evidence on leadership from nonindustrial and non-Western populuations. Shaman leaders, with their abilities to provide valuable benefits and impose severe costs, can play an outsized role in nonindustrial populations in numerous domains [@winkelman1987cross]. The WEIRD notion of secular nationalism, which separates the religious and temporal spheres [@juergensmeyer1993], can perhaps blind scholars raised in this tradition from recognizing how weird this separation is.

@@

In this data set (originally constructred by @garfield_evolutionary_2019) the number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one dimension specific to the current study.

@@

We investigated if evidence for shamans and other leaders with supernatural qualities varied by continental region and subsistence strategy. AIC values indicated that only region improved model fit. We did not include group context because these traits were predominantely coded in religious contexts. We also did not include leader sex because there were no text records of exclusively female leaders with evidence for shamanism (see results section).

>>>>

<!-- TITLE OPTIONS -->

<!-- * Leadership across human societies -->
<!-- * Human leadership: the determinants of leaders across cultures -->
<!-- * Leaders and followers across human societies -->
<!-- * Leadership across cultures and contexts -->
<!-- * The determinants of leadership across cultures and contexts -->
<!-- * The determinants of human leadership across societies -->

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
<!-- Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group context. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2020; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2020; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2020; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020]. -->

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, community leadership is typically informal and emerges in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015]. Leaders also play important roles in ritual and religious contexts [@singh_cultural_2017; @winkelman_crosscultural_2020], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Leadership in between-group conflict and cooperation is also common across nonindustrial societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community, and institutionalized managerial roles are associated with sedentarization, defensible resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of resources creates increased opportunities for resource management and is often associated with increased territoriality and inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

<!-- ASCRIBED THINGS -->
In the context of institutionalized social stratification, such as among chiefdoms, community leadership is often highly influenced by heredity and class structures [@earle_how_1997; @stanish_evolution_2004; @redmond_introduction_1998]. Managerial elites maintain influence over community decisions and are endowed with privileged social status [@stanish_evolution_2010]. Within state-level societies, including nonindustrial kingdoms, leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize political influence and maintain control through a variety of mechanisms including military force, ideologies, and control of information [@bodley_cultural_2011].

@@

Subsistence strategy, group structure and continental region were confounded. Hunter-gatherer text records often referred to residential and kin leadership, but rarely to supracommunity leadership; agriculturalist text records rarely referred to residential leadership but often referred to supracommunity leadership. Community-level political groups were frequent across all subsistence types. African text records were predominantly from horticulturalists whereas North American text records were predominantly from hunter-gatherers See Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence), and Table S\@ref(tab:fourfactorsumm).

@@

There were no text records that described shamans or leaders with supernatural qualities who were exclusively female. However, there were `r sum(df_shaman$demo_sex == 'both' & df_shaman$shamanism)` text records describing such leaders of both sexes. There was also more evidence of shamans and leaders with supernatural qualities in the Americas compared to other regions. See Figure S\@ref(fig:shamanregion).

@@

The leadership features we derived by clustering leadership dimensions were represented across most cultures. Leaders performed three broad functions: providing prosocial services, strategizing, and organizing collective actions. Regarding characteristic traits, leaders possessed some combination of individual competencies, many of which are prosocial, conformed to cultural norms, and received various benefits (Figure \@ref(fig:featuressupportplot)).

@@

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (Figure \@ref(fig:varsupportcb)). Nevertheless, for most dimensions of benefits and costs, there was more evidence of benefits than costs (Figure \@ref(fig:benefitcostratio)), and the same was true averaging across all benefit/cost dimensions ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$). Additionally, several positive leader qualities indicating social, reproductive, and material success clustered together (bottom cluster Figure \@ref(fig:cluster)A). These results are consistent with much theoretical literature that links leader benefits to their capacities to provide prosocial benefits (e.g., resolving conflicts, organizing collective actions, and generosity) and/or impose social costs (e.g., punishing and aggressiveness) [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013; @hooper_theory_2010; @gavrilets_solution_2014; @price_evolution_2014; @smith_leadership_2016].

@@

The potential evolutionary theoretical importance of shaman leaders (which in our data includes all leaders with supernatural qualities), highlights the importance of evidence on leadership from nonindustrial and non-Western populations. Shaman leaders, with their abilities to provide valuable benefits and impose severe costs, can play an outsized role in nonindustrial populations in numerous domains [@winkelman1987cross]. The WEIRD notion of secular nationalism, which separates the religious and temporal spheres [@juergensmeyer1993], can perhaps blind scholars raised in this tradition from recognizing how weird this separation is.

@@

In this data set (originally constructed by @garfield_evolutionary_2019) the number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one dimension specific to the current study.

@@

We investigated if evidence for shamans and other leaders with supernatural qualities varied by continental region and subsistence strategy. AIC values indicated that only region improved model fit. We did not include group context because these traits were predominantly coded in religious contexts. We also did not include leader sex because there were no text records of exclusively female leaders with evidence for shamanism (see results section).
